Reproducible Reseach - Peer Assignment #1
========================================================

##  Load and Process Data

```{r load, echo=TRUE}
#   Load data into R.
##  Set working directory
setwd("C:/Users/mmacrae/Coursera/5. Reproducible Research/Assignment #1")

##  Load activity data from CSV file into data frame.
activity <- read.csv('activity.csv')

#   Processing code converts measured data to analytic data.
##  Transform date from factor to date variable.
activity <- transform(activity, date=as.Date(date))

##  Combine date and time into a POSIXlt object in the data frame.
activity$datetime <- strptime(
    paste(format(activity$date, "%Y-%m-%s"), 
          sprintf("%04d", activity$interval)),
    "%Y-%m-%s %H%M"
)
```

##  Raw Data
### Calculate Daily Number of Steps

```{r data.raw}
#   What is the mean total number of steps taken per day?
##  Summarize number of steps per day.
steps.per.day <- with(activity, tapply(steps, as.Date(date), sum, na.rm=TRUE))

##  Plot histogram of the total number of steps taken each day.
hist(steps.per.day, col="blue",
     main="Daily Number of Steps", xlab="Daily Number of Steps")

##  Calculate and report the mean and median total number of steps taken per day.
mean.raw <- mean(steps.per.day)
median.raw <- median(steps.per.day)
```

The mean number of steps taken per day is `r mean.raw`.
The median number of steps taken per day is `r median.raw`.

### Average Daily Pattern

```{r pattern.raw}
##  Calculate average number of steps taken by interval.
steps.avg <- with(activity, tapply(steps, interval, mean, na.rm=TRUE))
steps.avg.time <- as.POSIXct(strptime(sprintf("%04d", as.numeric(names(steps.avg))), "%H%M"))

##  Plot time-series graph of the average number of steps taken in each 5-minute interval.
plot(steps.avg.time, steps.avg, type="l", col="blue", 
     main="Daily Activity Pattern", xaxt="n",
     xlab="Interval", ylab="Average Number of Steps")
axis.intervals <- seq(from=strptime("0000","%H%M"), to=strptime("2400","%H%M"), length.out=9)
axis(side=1,at=c(axis.intervals),
     labels=format(axis.intervals,format="%H:%M"))

##  Which five-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
interval.max <- strptime(sprintf("%04d",as.numeric(names(which.max(steps.avg)))),"%H%M")
interval.max.text <- with(interval.max,paste0(sprintf("%02d",hour),":",sprintf("%02d",min)))
abline(v=as.POSIXct(interval.max),lty=2,col="red")
```

The maximum number of steps, on average across all days in the dataset, occurs at `r interval.max.text`.

##  Clean Data
### Fill Missing Values

```{r n.missing, echo=TRUE}
n.missing <- sum(is.na(activity$steps))
```

There are `r n.missing` missing values in the raw data. We fill each missing value with the average number of steps taken in that hour across all days.

```{r data.clean}
## Copy activity data set to clean data step.
activity.clean <- activity

##  Fill in all missing values with average of the five-minute interval.
activity.clean$steps[is.na(activity.clean$steps)] <- 
    steps.avg[
        match(activity$interval[is.na(activity$steps)], names(steps.avg))
        ]
```

### Calculate Daily Number of Steps

```{r hist.clean}
##  Plot histogram of the total number of steps taken each day.
steps.per.day.clean <- with(activity.clean,tapply(steps,as.Date(date),sum,na.rm=TRUE))
hist(steps.per.day.clean, col="blue",
     main="Daily Number of Steps", xlab="Daily Number of Steps")

##  Calculate and report the mean and median total number of steps taken per day.
mean.clean <- mean(steps.per.day.clean)
median.clean <- median(steps.per.day.clean)
```

The mean number of steps taken per day is `r format(mean.clean,digits=5)`.
The median number of steps taken per day is `r format(median.clean,digits=5)`.

##  Differences in Activity Patterns Between Weekdays and Weekends

```{r activity.by.wday}
##  The $wday element in the POSITlt object returns an integer representing the day of week.
##  Define a vector of string objects indexed to $wday to group days into weekends and weekdays.
##  Sunday corresponds to one, and Saturday corresponds with seven.
##  Note: This simplified analysis fails to capture holidays.
wday.group <- factor(c("Weekend", rep("Weekday",5), "Weekend"))

##  In the $wday element of a POSIXlt object, Sunday corresponds to zero and Saturday corresponds with six.
##  Introduce a +1 adder to convert the $wday object to index the wday.group array.
activity.clean$wday <- wday.group[as.POSIXlt(activity.clean$datetime)$wday + 1]

##  Calculate the average steps by weekday/end and time interval.
steps.avg.by.wday <- with(activity.clean, sapply(split(steps, list(wday, interval)), mean))
x <- data.frame(steps.avg.by.wday,t(matrix(unlist(strsplit(names(steps.avg.by.wday), "[.]")),nrow=2)))
dimnames(x)[[2]] <- c("steps.avg","wday","interval")
x <- transform(x,interval=as.POSIXct(strptime(sprintf("%04d", as.numeric(as.character(interval))),"%H%M")))

##  Prepare 2x1 grid of plots.
par(mfcol=c(2,1))
axis.intervals <- seq(from=strptime("0000","%H%M"), to=strptime("2400","%H%M"), length.out=9)

#  Separate x into distinct weekend and weekday data frames.
weekend <- x[x$wday=="Weekend",]
weekday <- x[x$wday=="Weekday",]

#  Plot weekend time-series step data.
plot(as.POSIXct(weekend$interval), as.numeric(weekend$steps.avg), type="l", col="blue", 
     main="Weekend Activity Pattern", xaxt="n",
     xlab="Interval", ylab="Average Number of Steps",
     xlim=c(min(axis.intervals), max(axis.intervals)),
     ylim=c(0, max(x$steps.avg)))
axis(side=1, at=c(axis.intervals), labels=format(axis.intervals,format="%H:%M"))
abline(v=as.POSIXct(weekend[which.max(weekend$steps.avg),]$interval),lty=2,col="red")

##  Plot weekday time-series step data.
plot(as.POSIXct(weekday$interval), as.numeric(weekday$steps.avg), type="l", col="blue", 
     main="Weekday Activity Pattern", xaxt="n",
     xlab="Interval", ylab="Average Number of Steps",
     xlim=c(min(axis.intervals), max(axis.intervals)),
     ylim=c(0, max(x$steps.avg)))
axis(side=1, at=c(axis.intervals), labels=format(axis.intervals,format="%H:%M"))
abline(v=as.POSIXct(weekday[which.max(weekday$steps.avg),]$interval),lty=2,col="red")
```
